Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory

### 2024-07-09 POPL/paper.tex
spelling

  We give this type theory a semantics in the category of grammars and prove a
  canonicity theorem showing that every term in a context corresponding to a
  fixed string is equal to a term in a canonical form encoding a parse tree of
  that string. Finally, we discuss the application to the design and
  implementation of correct-by-contruction parsers.
\end{abstract}

@@

We say that a grammar $G$ associates to every string $w$ the set\footnote{We
  make little comment as to which foundations are used to encode $\Set$. Our
  construction is polymorphic in choicce of a proper class of small sets, a
  Grothendieck universe, a type theoretic universe, or any similar foundation.} $G w$ of parse
trees showing that $w$ matches $G$. This definition of formal grammar serves as
a mathematical specification for a parser --- taking on the same role as
languages do for recognizers. In this sense, our new characterization of
grammars is just a proof-relevant generalization of the existing notion of
formal language.

@@

The category-theoretic framework can be further used to describe and compare
different notions of formal grammar. For instance, Chomskyan generative
grammars, semi-Thue systems, Montague grammars, and so on are all distinct syntactic
presentations of the same underlying idea of an abstract specification for
parsing. Each of these is an instance of a general \emph{notion of formal
  grammar} --- a category paired with a functor into
$\Set^{\String}$. That is, they may all be interpreted in our category of grammars. In this sense, $\Set^{\String}$ provides a universal setting to reason about parsing.
The slice 2-category over $\Set^{\String}$ then provides the appropriate expressivity to prove equivalences between grammar formalism, such as the equivalence between semi-Thue systems and Chomsky's unresticted grammars.

@@

\begin{itemize}
  \item A universal semantic notion of formal grammar given by the category $\Set^{\String}$
  \item Categorical structures that capture the concepts integral to formal grammar theory, and that are notably modeled by $\Set^{\String}$
  \item A dependent
    linear/non-linear (LNL) type theory to syntactically capture the above categorical structures
  \item An implementation of the language, which is emedded shallowly in Agda
\end{itemize}

\section{Categorification of Formal Grammars}
\label{sec:categorify}
\steven{Idk if this section has the right title, but I wanted to change it if we are adding ``chomsky categories''}

Kleene algebras are an important tool in the theory of regular
languages. More broadly, they serve as a theoretical substrate to
studying various kinds of formal languages. Formally, they are a tuple
$(A, +, \cdot, (-)^*, 1, 0)$, where $A$ is a set, $+$ and $\cdot$
are binary operations over $A$, $(-)^*$ is a function over $A$, and
$1$ and $0$ are constants. These structures satisfy the axoims depicted
in Figure~\ref{fig:axioms}.

@@

\begin{definition}
  A Kleene category is a distributive monoidal category $\cat{K}$
  such that for every objects $A$ and $B$, the endofunctors $F_{A, B}
  = B + A \otimes X$ and $G_{A, B} = B + X \otimes A$ have initial
  algebras (denoted $\mu X.\, F_{A, B}(X)$) such that $B \otimes (\mu
  X.\, F_{A, 1}) \cong \mu X.\, F_{A, B}(X)$ and the analogue isomorphism
  for $G_{A,B}$ also holds.
\end{definition}

@@

\steven{Maybe there is more to say here, or perhaps later, that the only thing
  to gain parsing-wise from a non-dependent non-linear fragment are the things that are
  definiable via $G$. Except that these are not interesting, as they are
  necessarily empty}

@@

\steven{Mention sepration logic at some point. In particular the separating conjunction and magic wand}

@@

The non-linear
subtheory corresponds to Martin-L\"of type theory (MLTT), which can be taken
as a foundation for mathematics. Dependence of linear terms on non-linear terms
of this form effectively allows us to bring any ordinary definitions from math
into scope as indicies and then define grammars relative to those indices. That
is all to say, the LNL dependent types in the style of Krishnaswami et al.\
allows us to recover the defintion of Aho's indexed grammars \cite{AhoIndexed} internal to our
type theory.
\steven{The above paragraph should maybe be put later. Also flows weirdly}

@@

The structrual judgements are shown in \cref{fig:structjdg}, the typing
well-formedness in \cref{fig:typewf}, and the intuitionistic typing rules in
\cref{fig:inttyping}. These are all just as they appear in
\cite{krishnaswami_integrating_2015}.

@@

$\Gamma$ reprsents a non-linear context, while $\Delta$ represents a
\emph{linear context} dependent on $\Gamma$. These linear contexts behave
substructurally. As stated earlier, they do are linear --- they do not obey
weakening or contraction --- because a character is exhausted once read by a
parsing procedure. Moreover, the characters in strings appear in the order in
which we read them. We do not have the freedom to freely permute characters,
therefore any type theory that is used to reason about formal grammars ought to
omit the structural rule of exchange as well. This means that every variable
within $\Delta$ must be used \emph{exactly once} and \emph{in order of
  occurrence}. Thus, we can think of the linear contexts as an ordered list of
limited resources. Once a resource is consumed, we cannot make reference to it
again. Variables in a linear context then act like building
blocks for constructing patterns over strings.

We give the base types and type constructors for linear terms. As the
interpretation of types as grammars in $\Set^{\String}$ serves as our intended
semantics, we simultaneously give the interpreation $\sem \cdot$ of the
semantics as grammars.

@@

The grammar semantics of $A \amp B$ denotes a pair of a parse of $A$ and a
parse of $B$ \emph{of the same string}. In this sense, $\amp$ acts a
conjunction between grammars --- meaning a string only matches $A \amp B$
if it simlutaneously matches each of $A$ and $B$.

>>>>

  We give this type theory a semantics in the category of grammars and prove a
  canonicity theorem showing that every term in a context corresponding to a
  fixed string is equal to a term in a canonical form encoding a parse tree of
  that string. Finally, we discuss the application to the design and
  implementation of correct-by-construction parsers.
\end{abstract}

@@

We say that a grammar $G$ associates to every string $w$ the set\footnote{We
  make little comment as to which foundations are used to encode $\Set$. Our
  construction is polymorphic in choice of a proper class of small sets, a
  Grothendieck universe, a type theoretic universe, or any similar foundation.} $G w$ of parse
trees showing that $w$ matches $G$. This definition of formal grammar serves as
a mathematical specification for a parser --- taking on the same role as
languages do for recognizers. In this sense, our new characterization of
grammars is just a proof-relevant generalization of the existing notion of
formal language.

@@

The category-theoretic framework can be further used to describe and compare
different notions of formal grammar. For instance, Chomskyan generative
grammars, semi-Thue systems, Montague grammars, and so on are all distinct syntactic
presentations of the same underlying idea of an abstract specification for
parsing. Each of these is an instance of a general \emph{notion of formal
  grammar} --- a category paired with a functor into
$\Set^{\String}$. That is, they may all be interpreted in our category of grammars. In this sense, $\Set^{\String}$ provides a universal setting to reason about parsing.
The slice 2-category over $\Set^{\String}$ then provides the appropriate expressivity to prove equivalences between grammar formalism, such as the equivalence between semi-Thue systems and Chomsky's unrestricted grammars.

@@

\begin{itemize}
  \item A universal semantic notion of formal grammar given by the category $\Set^{\String}$
  \item Categorical structures that capture the concepts integral to formal grammar theory, and that are notably modeled by $\Set^{\String}$
  \item A dependent
    linear/non-linear (LNL) type theory to syntactically capture the above categorical structures
  \item An implementation of the language, which is embedded shallowly in Agda
\end{itemize}

\section{Categorification of Formal Grammars}
\label{sec:categorify}
\steven{Idk if this section has the right title, but I wanted to change it if we are adding ``Chomsky categories''}

Kleene algebras are an important tool in the theory of regular
languages. More broadly, they serve as a theoretical substrate to
studying various kinds of formal languages. Formally, they are a tuple
$(A, +, \cdot, (-)^*, 1, 0)$, where $A$ is a set, $+$ and $\cdot$
are binary operations over $A$, $(-)^*$ is a function over $A$, and
$1$ and $0$ are constants. These structures satisfy the axioms depicted
in Figure~\ref{fig:axioms}.

@@

\begin{definition}
  A Kleene category is a distributive monoidal category $\cat{K}$
  such that for every objects $A$ and $B$, the endofunctors $F_{A, B}
  = B + A \otimes X$ and $G_{A, B} = B + X \otimes A$ have initial
  algebras (denoted $\mu X.\, F_{A, B}(X)$) such that $B \otimes (\mu
  X.\, F_{A, 1}) \cong \mu X.\, F_{A, B}(X)$ and the analogous isomorphism
  for $G_{A,B}$ also holds.
\end{definition}

@@

\steven{Maybe there is more to say here, or perhaps later, that the only thing
  to gain parsing-wise from a non-dependent non-linear fragment are the things that are
  definable via $G$. Except that these are not interesting, as they are
  necessarily empty}

@@

\steven{Mention separation logic at some point. In particular the separating conjunction and magic wand}

@@

The non-linear
subtheory corresponds to Martin-L\"of type theory (MLTT), which can be taken
as a foundation for mathematics. Dependence of linear terms on non-linear terms
of this form effectively allows us to bring any ordinary definitions from math
into scope as indices and then define grammars relative to those indices. That
is all to say, the LNL dependent types in the style of Krishnaswami et al.\
allows us to recover the definition of Aho's indexed grammars \cite{AhoIndexed} internal to our
type theory.
\steven{The above paragraph should maybe be put later. Also flows weirdly}

@@

The structural judgments are shown in \cref{fig:structjdg}, the typing
well-formedness in \cref{fig:typewf}, and the intuitionistic typing rules in
\cref{fig:inttyping}. These are all just as they appear in
\cite{krishnaswami_integrating_2015}.

@@

$\Gamma$ represents a non-linear context, while $\Delta$ represents a
\emph{linear context} dependent on $\Gamma$. These linear contexts behave
substructurally. As stated earlier, they do are linear --- they do not obey
weakening or contraction --- because a character is exhausted once read by a
parsing procedure. Moreover, the characters in strings appear in the order in
which we read them. We do not have the freedom to freely permute characters,
therefore any type theory that is used to reason about formal grammars ought to
omit the structural rule of exchange as well. This means that every variable
within $\Delta$ must be used \emph{exactly once} and \emph{in order of
  occurrence}. Thus, we can think of the linear contexts as an ordered list of
limited resources. Once a resource is consumed, we cannot make reference to it
again. Variables in a linear context then act like building
blocks for constructing patterns over strings.

We give the base types and type constructors for linear terms. As the
interpretation of types as grammars in $\Set^{\String}$ serves as our intended
semantics, we simultaneously give the interpretation $\sem \cdot$ of the
semantics as grammars.

@@

The grammar semantics of $A \amp B$ denotes a pair of a parse of $A$ and a
parse of $B$ \emph{of the same string}. In this sense, $\amp$ acts a
conjunction between grammars --- meaning a string only matches $A \amp B$
if it simultaneously matches each of $A$ and $B$.
